---
title: 'Homework 04: Tidy data and joins'
output: github_document
---
## Tidy Cheatsheet (Data Reshaping Prompts)
This cheatsheet aims to explore several basic tools in tidyr that will help organize our data in a consistent way in R. Tidyr is a member of the core tidyverse package. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(nycflights13))
```

## Gather() and Spread()

Let's use _gapminder_ data to explore these tools

### Gather()

I will use _t1_ dataset to demonstrate `gather()`:
```{r}
t1=gapminder %>%
  filter(continent == "Asia" & year == "2002") %>%
  select(year, country, pop, lifeExp) %>%
  group_by(year) 
t1
```

I will use `gather()` to rearrange values from population and life expectancy in same column and use a key to differentiate them.

```{r}
t1 %>% 
  gather(pop, lifeExp, key=parameter, value=value) %>% 
  arrange(year, country) 
```


### Spread()

Using another data, _t2_, as an example: a tibble with 24 rows: 2 per year, giving the country with both the lowest and highest life expectancy (in Asia)
```{r}
t2=gapminder %>%
  filter(continent == "Asia") %>%
  select(year, country, lifeExp) %>%
  group_by(year) %>%
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2) %>% 
  arrange(year) %>%
  print(n = Inf)
```

Now I want to turn the table into one row per year using `spread()`

```{r}
t2 %>% 
  spread(key=country, value=lifeExp)

```

### Unite()

We can use `unite() `function to combine two columns into one. Using _t2_ as an example, assume we want to combine country and lifeExp into one column:

```{r}
t2 %>% 
  unite(lifeInCountry, country, lifeExp, sep = ", ")

```

Now we have one column contain each country and its life expectancy. I found this function is very helpful in organizing summary of a regression model.

## Cheatsheet for dyplr join functions (Join Prompts)

I will use the _nycflight13_ package to explore different join functions in dyplr. _nycflight13_ contains several tibbles that are related to flights. In this deomstration, I will only use follow two datasets: 
```{r}
flights
airlines
```


We can see that _flights_ contains too many variables and observations, so let's narrow the dataset before using it for demonstration.

```{r}
flight2 = flights %>% 
  filter(day=="6" & hour=="8" & dest == "IAH") %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
  
flight2 ## I will use this data for following demonstration
```

### Inner_join
Inner_join(x, y) is the simplest type of join. It matches pairs of observations whenever their keys are equal, i.e. it returns all columns from x and y. So it is a mutating join.

Let's creat a dataset to join _flight2_
```{r}
dat1=tribble(
  ~tailnum, ~month,
  "N76502", 1, 
  "N14731", 2,
  "N438UA", 12
  )
dat1
```

Now let's join dat1 to _flight2_.
```{r}
flight2 %>% 
  inner_join(dat1, by="month")
```
We can see that results only return to records when keys are equal.

### Left_join
left_join(x, y) joins matching rows from y to x. Similar, `right_join()` matches rows from x to y.

Let's try left_join and use flight2 for demonstration:
```{r}
flight2 %>% 
  left_join(airlines, by="carrier")
```
The result of left join _airlines_ to _flight2_ return an additional variable: name. I am wondering if any basic R functions could achieve the same outcome?

### Comparing with R basic function `match()` 
Same results as above left join can also be obtained using `mutate()` combining wtih with R basic function `match()`

```{r}
flight2 %>% 
  mutate(name=airlines$name[match(carrier, airlines$carrier)])
```

It generates exactly same result! But this way is harder to think through and apply when there are multiple variables. More importantly, the code itself require extra efforts to understand its intention.

### Full_join()

full_join(x, y): Return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing. This type of join could be very helpful when dealing with real data, which we want to keep all records.

Let's use previous example from left_join again, see what happen:
```{r}
flight2 %>% 
  full_join(airlines, by="carrier")
```

Using full_join we now have 30 records with later half of them filled by NA, compared to left_join where we only have 15 records that matches _flight2_.



